\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage{microtype}
\usepackage{amsmath,amssymb,amsthm,mathtools}
\usepackage{booktabs}
\usepackage[numbers,sort&compress]{natbib}
\usepackage{hyperref}

\hypersetup{
  colorlinks=true,
  citecolor=blue,
  linkcolor=blue,
  urlcolor=blue
}

\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\Q}{\mathbb{Q}}
\newcommand{\LLL}{\operatorname{LLL}}
\newcommand{\bits}{\operatorname{bits}}
\setlength{\emergencystretch}{3em}

\title{Certified Exact Integer LLL in Lean 4:\\
Explicit Inverses, Total Rank Handling, and Profiled Costs}
\author{Junye Ji}
\date{Research version 0.1.0, July 2026}

\begin{document}

\maketitle

\begin{abstract}
The Lenstra--Lenstra--Lov\'asz algorithm replaces an integer lattice basis by
a shorter, nearly orthogonal basis while preserving the generated lattice.
We present an executable Lean 4 development with a fixed row-vector
convention, exact parameters $\delta=3/4$ and $\eta=1/2$, and no
floating-point arithmetic.  A source-pinned verified dense reducer supplies
termination and reducedness.  A synchronized trace replays every size
reduction and adjacent swap as an elementary integer row operation, producing
$U$, a chosen $U^{-1}$, proofs of both inverse identities, and $UB=B'$.  A
verified Hermite rank decomposition extends the independent-row kernel to
empty, zero, rectangular, and rank-deficient inputs; a column interface is
obtained only by transposition.  A profiled entry records the proved
termination fuel, exact event telemetry, and a running height covering every
transformed basis and row-add multiplier.  We prove an event bound and compose
it with the project's verified sign-magnitude arithmetic charges.  The result
is deliberately an a-posteriori, profile-dependent reference budget, not an
input-only polynomial complexity theorem.  The independent artifact freezes
114 API entries, audits 20 theorem roots, and records raw one-warmup,
seven-measurement observations for four deterministic cases.
\end{abstract}

\section{Introduction}

LLL reduction is both a central lattice algorithm and a common preprocessing
step for exact matrix algorithms~\citep{lll1982,cohen1993}.  A formal
implementation must make several conventions explicit.  Are vectors stored
as rows or columns?  Which size-reduction interval is accepted at a tie?  Is
linear independence an unchecked precondition?  Does a reported unimodular
transform include an actual inverse, or merely a determinant claim?  What
happens to dependent rows?  Finally, which arithmetic is included in a cost
statement?

The facade
\begin{center}
  \code{NormalForms.Research.LLL}
\end{center}
answers these questions without changing the stable normal-forms API.  Lean 4
and mathlib supply the ambient exact matrix theory~\citep{demoura2021lean4,
mathlib2020}.  The contributions are:
\begin{itemize}
  \item a fixed exact LLL predicate for integer row bases;
  \item a total independent-row reducer whose transform and chosen inverse
        are accumulated from the executed elementary operations;
  \item a total rank-deficient wrapper with a reduced nonzero prefix and a
        literal zero tail;
  \item a transpose-only column adapter, avoiding a competing convention;
  \item an execution profile with a proved fuel-dependent event bound and
        intermediate coefficient evidence; and
  \item an explicitly scoped binary reference budget, benchmark, audit,
        manuscript, and pinned artifact.
\end{itemize}

\section{Exact row-basis semantics}

Let $B\in\Z^{m\times n}$, with row $b_i$.  The implementation casts rows to
$\Q^n$ and computes exact Gram--Schmidt vectors $b_i^*$ without square roots.
Writing
\[
  \mu_{ij}=\frac{\langle b_i,b_j^*\rangle}
                   {\langle b_j^*,b_j^*\rangle},
\]
the public predicate \code{IsLLLReduced B} contains three fields.  First,
every Gram determinant is positive, so the rows are independent.  Second,
for $j<i$,
\[
  |\mu_{ij}|\le\eta=\frac12.
\]
Third, every adjacent pair satisfies
\[
  \delta\lVert b_i^*\rVert^2
  \le
  \lVert b_{i+1}^*\rVert^2+
  \mu_{i+1,i}^2\lVert b_i^*\rVert^2,
  \qquad \delta=\frac34.
\]
The executable \code{isLLLReducedB} uses the same exact integer-backed data,
and \code{isLLLReducedB\_sound} promotes acceptance to the proposition.  No
floating point tolerance occurs in either path.

The full-rank input contract stores executable independence, rather than a
rational right inverse or determinant selected by choice.  The result type
contains
\[
  U,\quad U^{-1},\quad B',\quad UB=B',\quad
  U^{-1}U=I,\quad UU^{-1}=I,
\]
together with exact reducedness of $B'$.

\section{Terminating backend and explicit transform trace}

The dense reducer is source-pinned through the Lake manifest.  Its integer
state stores the live basis, scaled Gram--Schmidt coefficients, and leading
Gram determinants.  Termination uses the classical product potential and the
lexicographic row index.  For initial state $s$, the executable fuel is
\[
  F(s)=(\operatorname{potential}(s)+1)(m+1).
\]
The imported theorem proves this fuel reaches the completed branch for an
independent basis; the zero-fuel fallback is therefore unreachable through
the public positive-row entry.

Our trace follows the same branch decisions and fold order.  If a size
reduction selects integer $q$, it applies the elementary operation
\[
  b_i\leftarrow b_i-qb_j.
\]
The forward elementary matrix is inverted by changing $q$ to $-q$.  An
adjacent row swap is self-inverse.  At each step the trace composes the new
certificate with the accumulated certificate.  Consequently the final
inverse is data obtained from execution, not a matrix inverse reconstructed
from a nonzero determinant.  A representation theorem proves that erasing
the certificate trace yields exactly the verified dense output.

The opaque public entry prevents recursive state and private helpers from
entering downstream native call graphs.  Its return type nevertheless exposes
all mathematical data and equations.

\section{Total rank-deficient and column interfaces}

LLL's independent-row kernel cannot directly accept dependent rows.  The
total row entry first computes a strong Hermite result
\[
  U_HB=H
\]
with explicit $U_H^{-1}$.  From the Hermite pivot rules we construct a profile
with strictly increasing pivot columns, positive pivots, zeros below each
pivot, and a literal zero-row tail.  For the leading $r$ rows, the selected
pivot minor is upper triangular and nonzero.  Cauchy--Binet then proves every
leading Gram determinant positive, which matches the dense backend's
executable independence predicate.

The independent prefix is reduced by the full-rank kernel.  Its transform is
lifted to
\[
  \widehat U=\operatorname{diag}(U_L,I_{m-r}).
\]
The lift has the blockwise inverse
$\operatorname{diag}(U_L^{-1},I_{m-r})$.  Composing certificates gives the
total transform $\widehat U U_H$ and its chosen inverse.  The result proves
that its first $r$ rows are LLL reduced and all later rows are zero.  This
covers $r=0$, empty dimensions, rectangular inputs, and arbitrary row rank.

For column vectors, the only public adapter applies the row algorithm to
$B^T$ and transposes the result.  Thus it returns $V$ with
\[
  BV=B'
\]
and both inverse identities.  No separately maintained column reduction can
drift from the fixed row convention.

\section{Profiled operation and coefficient evidence}

The profiled full-rank entry returns the same strong contract and an execution
record.  Its counter separates successful size reductions, swaps, and the
associated dense-state refreshes.  A sweep at row $k$ examines exactly $k$
earlier positions, hence performs at most $m$ successful reductions.  One
outer transition performs at most one swap.  Because every successful row
operation also records one refresh, induction over the structurally bounded
loop proves
\[
  \operatorname{events}
  \le F(s)(2m+2).
\]

The trace also maintains a monotone maximum.  Initially it is the largest
absolute input entry.  At a reduction it is enlarged by both the next basis
height and $|q|$; at a swap it is enlarged by the next basis height.  Therefore
the stored number covers every transformed basis and every row-add multiplier
encountered by the trace.  Monotonicity of binary length yields
\[
  \bits(B')\le\bits(H_{\mathrm{observed}}).
\]
This is genuine intermediate evidence, but the observed height remains an
output of the profile rather than an input-only formula.

\section{Binary reference budget and scope}

The shared bit-cost library represents integers by a sign and canonical binary
magnitude.  It implements costed carry/borrow addition, schoolbook
multiplication, and long division with remainder, and proves their values equal
the corresponding integer operations.  If $h$ is the observed coefficient
height, the LLL profile uses the working width
\[
  W(h)=2\bits(h)+3.
\]
The common scalar charge is the sum of the verified upper charges for one
addition, one multiplication, two divisions, and a linear comparison.  The
formal lemmas show each primitive call at width $W(h)$ fits this charge.
Multiplying by a conservative number of scalar positions per event and the
proved event budget produces \code{lllKernelBitOperationBound}; the profile's
\code{bitOperationCost\_le} theorem closes the composition.

This statement has a narrow, testable meaning.  It prices a specified
sign-magnitude reference model at the trace's observed basis-and-multiplier
height.  It is not an input-only polynomial LLL complexity theorem.  It does
not include HNF rank decomposition, multiplication of accumulated transform
matrices, Lean allocation, garbage collection, or compiler behavior.  Native
wall time is reported only as observational evidence.  These exclusions are
part of the API documentation and artifact validation, rather than left as an
informal footnote.

\section{Evaluation}

The deterministic corpus has two profiled independent-row cases and two total
interface cases.  Table~\ref{tab:cases} records the theorem-facing metrics.

\begin{table}[ht]
\centering
\caption{Deterministic LLL corpus. The cost columns apply only to profiled
full-rank cases.}
\label{tab:cases}
\begin{tabular}{lrrrrrr}
\toprule
Case & Rank & Fuel & Events & Bound & Int. bits & Cost / bound \\
\midrule
Gaussian $2\times2$ & 2 & 9 & 4 & 54 & 1 & 182976 / 2470176 \\
Dense $3\times3$ & 3 & 1252 & 16 & 10016 & 3 & 4230720 / 2648430720 \\
Dependent rows & 1 & -- & -- & -- & -- & -- \\
Column adapter & 1 & -- & -- & -- & -- & -- \\
\bottomrule
\end{tabular}
\end{table}

Every case checks the transformation equation and both inverse products.  The
full-rank cases additionally check exact reducedness and both profile bounds;
the dependent cases check rank, reduced prefix, and zero tail.  On the
captured host, median wall times after one warmup and seven measurements were
$0.02$ s for the Gaussian case, $0.20$ s for the dense case, $0.06$ s for the
dependent row wrapper, $0.06$ s for the column adapter, and $0.27$ s end to
end.  Raw observations, inclusive-quartile summaries, RSS, hardware details,
source state, and binary hash are preserved in JSON and CSV.  No portable
wall-time threshold is inferred from them.

\section{Trust and reproducibility}

The research facade is absent from the frozen core facade; a negative import
harness checks the exact unknown-identifier diagnostic.  A 114-entry
compilation manifest freezes public names and types.  The axiom audit starts
from 20 theorem roots and records the complete dependency closure as JSON.
Its union is exactly \code{propext}, \code{Quot.sound}, and
\code{Classical.choice}; it rejects project axioms, \code{sorryAx}, and
native/compiler trust.

Small benchmark rank witnesses use native decision inside the executable
only.  Those fixtures are not public theorem roots and native success never
constructs a library result.  The library's semantic and cost theorems are
kernel checked.  A pinned Debian container installs the hash-verified Lean
4.32.0 archive, restores the locked Lake dependencies, and runs the same
release gate.  The gate performs the build, complete test suite, direct
execution guards, facade isolation, audit, baseline validation, and paper
build.  It prepares local release material only.

\section{Conclusion}

The development turns the usual LLL output convention into a total, explicit
contract: the basis orientation and parameters are fixed; the transform has a
chosen inverse with both identities; dependent rows are moved to a proved zero
tail; and column use is only a transpose adapter.  The profiled path separates
semantic correctness, event counts, intermediate coefficient evidence, and a
specified binary reference price.  Its deliberately limited cost theorem
provides a reproducible foundation for a future input-only polynomial analysis
without overstating what the present implementation proves.

\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
