We formalize certified computation of homology for bounded chain complexes of
finite free abelian groups in Lean 4.  Boundary maps are explicit integer
matrices with the fixed column-vector convention
$d_{n+1}\colon C_{n+1}\to C_n$.  A first strong Smith calculation on $d_n$
provides a chosen right transformation and its inverse.  Its tail columns
form an explicit integral basis of $\ker d_n$, while the chain condition
proves that the leading coordinates of $V^{-1}d_{n+1}$ vanish.  Restricting
to the tail produces a presentation matrix for homology.  A second strong
Smith calculation then yields the torsion factors, Betti number, and an
explicit additive equivalence with cyclic and free summands.  The
implementation connects this semantic construction to a fully executable
matrix projection and validates the circle, torus, real projective plane,
and a finite cellular example.  A finite normalized simplicial frontend
derives alternating boundary matrices from explicitly ordered nondegenerate
simplices, retaining repeated-face cancellation and erasing degenerate faces.
Every explicit matrix complex is realized as a mathlib homological complex,
and an explicit degree-zero comparison identifies the intrinsic quotient
with categorical homology.  Given a certified realization of the finite face
data by a genuine simplicial set, chain isomorphisms transport the result to
mathlib's nondegenerate normalized chains and normalized Moore complex.
An independent facade, native test, axiom audit, release gate, and artifact
preserve the stable normal-form core.
