\section{Execution, audit, and artifact}

The executable regression covers every represented degree of four finite
cellular complexes:

\begin{center}
\begin{tabular}{lll}
\toprule
Complex & Degree & Computed homology\\
\midrule
Circle & 0, 1 & $\Z,\ \Z$\\
Torus & 0, 1, 2 & $\Z,\ \Z^2,\ \Z$\\
Real projective plane & 0, 1, 2 & $\Z,\ \Z/2,\ 0$\\
Degree-six attachment & 0, 1, 2 & $\Z,\ \Z\times\Z/6,\ 0$\\
\bottomrule
\end{tabular}
\end{center}

The native executable checks twenty complete summaries, including chain,
outgoing, and kernel ranks; all invariant factors; filtered torsion factors;
and Betti numbers.  The eight added summaries cover the minimal simplicial
circle, collapsed-boundary sphere, and filled triangle. Tests also check
their exact face signs, degeneracy marker, boundary matrices, chain equation,
and inverse identities from both Smith calculations. Per-case and total
nanoseconds are recorded as observational diagnostics, not as portable
performance thresholds.

The application axiom audit covers 37 roots: the chain condition, simplicial
boundary derivation and conversion, both strong Smith results, zero-prefix
theorem, kernel equivalence, quotient equivalence, decomposition,
theorem/runtime bridges, categorical homology comparison, and normalized
Moore transports.  It observes only
\code{propext}, \code{Quot.sound}, and \code{Classical.choice}.  Choice is
admitted in the semantic basis and quotient layer.  No project axiom,
\code{sorryAx}, or compiler/native trust axiom is allowed.

The independent release gate builds the homology facade and its 53-declaration
public API freeze, runs the full repository test suite, executes the native cases,
validates the exact audit allowlist, and rebuilds this paper.  A pinned
container repeats the code checks without TeX and without FLINT.  Native
success is only a regression report; the Lean kernel checks the mathematical
claims.
