We report the version 1 core of a \Lean{} formalization of Hermite and Smith
normal forms over Euclidean domains. The recursive kernels operate on
\code{Fin}-indexed matrices, while every arbitrary finite-index entry point
stores the exact row and column enumeration chosen by the caller. Results carry
explicit inverse matrices and both inverse identities, rather than recovering
inverses from determinants. We prove HNF correctness and fixed-index
uniqueness, SNF correctness and uniqueness, and invariance of the canonical
\code{Fin} Smith matrix and invariant factors under changes of indexing.
\code{SmithData} records the complete normalized diagonal and zero tail;
\code{SmithSignature} records its intrinsic associate factors. Determinantal
ideals recover the rank and factors and prove invariance under certified
transforms and reindexing. A canonical wrapper compares this signature with
\mathlib{}'s PID Smith witnesses.

A constructive Euclidean-operation seam supplies execution over
\(\mathbb Z\) and \(\mathbb Q[X]\), including direct polynomial long division
and extended gcd, while Smith termination uses a pivot measure instead of
factorization. Column-oriented finite-free presentations define cokernels,
Fitting ideals, Smith decompositions, and a bridge to
\code{Module.Presentation}. The external path parses and checks untrusted
certificates, emits chunked \code{decide_cbv} obligations, and rechecks
optional FLINT-generated data in the Lean kernel. Process and FFI adapters
share one C implementation and raw schema; seeded differential tests compare
only canonical invariant factors. A reproducible profile separates internal
generation, external generation, native checking, strict kernel checking, and
total import costs. Exact type tests and separate axiom audits freeze the
public API and trusted execution boundary; a pinned artifact profile rebuilds
the complete release gate.
