\section{Future Work: Towards Smith Normal Form and the PID Bridge}
\label{sec:future}

The companion repository already contains a stable public interface for
\SNF{}, together with a genuine specification layer and certificate packaging
boundary. The module
\module{NormalForms.Matrix.Smith.Basic} currently defines:

\begin{itemize}
  \item \code{IsSmithNormalForm}, which is a real diagonal-shape specification
    phrased over arbitrary finite index types;
  \item \code{smithInvariantFactors} and \code{smithColumnSpan}, which expose
    the current public readout layer;
  \item \code{SNFResult}, which packages matrices \(U\), \(S\), \(V\), the
    equation \(U A V = S\), and a Smith-form witness;
  \item \code{SNFResult.ofCertificate} and \code{SNFResult.toCertificate},
    which define the current public certificate/result boundary;
  \item \code{smithNormalForm}, whose signature already carries
    \code{[CanonicalMod R]} but whose current implementation returns
    \code{none}.
\end{itemize}

This paper therefore makes \emph{no} claim about executable Smith normal form, no
claim about Smith-form correctness or uniqueness, and no claim about the final
classification theorem for finitely generated abelian groups. Those topics remain
future work.

Internally, however, the repository has already moved past a pure API freeze.
The same module now contains proof-carrying pivot states together with a
verified same-size lead-reduction foundation: raw row/column clearing steps,
pure-data first-column and first-row clearing loops, external invariance
theorems, and a wrapper \code{clearLeadByPivot} that reassembles the resulting
state. This is meant as the base layer for the eventual executable \SNF{}
kernel, not as a completed Smith algorithm.

The current \SNF{} API nevertheless records two design decisions worth making
explicit. First, \code{SNFResult} deliberately does \emph{not} include
unimodularity fields for \(U\) and \(V\). The HNF experience showed that
certificate packaging matters; freezing the public boundary before the proofs are
written should not over-commit to a final proof object layout. Second,
\code{CanonicalMod} appears on the executable entrypoint \code{smithNormalForm},
not on the abstract predicate \code{IsSmithNormalForm}. This mirrors the HNF
architecture: canonical remainder assumptions belong on executable exactness, not
on the abstract shape of the normal form itself.

There is also a deliberate elaboration boundary in the current design. The
public Smith wrappers reindex arbitrary finite types through \code{Fintype.equivFin},
but the repository intentionally avoids aggressive public \code{simp} lemmas that
collapse this bridge on concrete \code{Fin}-indexed matrices. In practice those
lemmas make elaboration unstable. Concrete diagonal-specification and
invariant-factor smoke tests therefore remain on the internal \code{Fin} layer,
while the public examples focus on stable packaging statements.

The same caution applies to the PID bridge. The dedicated bridge module currently
contains only a placeholder agreement predicate together with a trivial theorem
named \code{executableSNFMatchesPID}. These declarations are markers
for the intended integration with \mathlib{}'s PID-side structural results, not
completed mathematics. A realistic next milestone is therefore:

\begin{itemize}
  \item extend the current lead-reduction foundation with nondivisible pivot
    improvement and a strict descent argument;
  \item connect that same-size layer to the outer recursive
    \code{smithNormalFormFin} kernel;
  \item prove public correctness, left/right unimodularity extraction, and
    Smith uniqueness with explicit two-sided certificates;
  \item only then relate the resulting invariant factors to the abstract PID
    theorems already available in \mathlib.
\end{itemize}
