\section{Execution, audit, and artifact}

The regression suite covers scalar and diagonal matrices, a nontrivial
coordinate swap, a size-two Jordan block, exact companion layouts, both
change-of-basis inverse identities, and dimension zero.  The independent API
test compiles the main entry points and exact result equations.  The
rational-canonical axiom audit observes only \code{propext},
\code{Quot.sound}, and \code{Classical.choice}.  Choice occurs in the semantic
quotient and basis layer; no project axiom, \code{sorryAx}, or native/compiler
trust axiom is admitted.

The native benchmark uses
\[
  A=\begin{pmatrix}1/2&1\\0&1/2\end{pmatrix}.
\]
Its factors have degrees $[0,2]$, and the maximum rational coefficient bit
length grows from 2 in the input to 3 in the output.  Table~\ref{tab:bench}
summarizes one warmup followed by seven observations on the host fingerprint
stored with the raw data.

\begin{table}[h]
\centering
\begin{tabular}{lrr}
\toprule
Stage & Median wall time & Median maximum RSS\\
\midrule
Invariant generation & 0.07 s & 72,728 KiB\\
Companion batch (1,000 checks) & 0.68 s & 72,844 KiB\\
End to end (one check) & 0.07 s & 72,896 KiB\\
\bottomrule
\end{tabular}
\caption{Native rational-canonical profile.  The isolated median companion
batch is 0.617 s, approximately 617 microseconds per structural check.}
\label{tab:bench}
\end{table}

Wall time is observational and is never a portable hosted-CI threshold.  The
authoritative JSON and CSV contain all observations, inclusive-quartile
summaries, source revision, and hardware data.  Operation, pivot, and gcd
counts remain explicitly unavailable; assigning guessed counts would blur
the boundary with the separate formal bit-cost project.

The artifact profile rebuilds the application facade, API tests, full
regression driver, native executable, audit, and fresh benchmark.  It does
not link FLINT.  Benchmark success is not used to construct a theorem; the
kernel-checked Smith, module, companion, and similarity proofs establish
correctness.
