\section{Introduction}

The rational canonical form classifies an endomorphism of a finite-dimensional
vector space over a field without choosing eigenvalues or splitting the
characteristic polynomial.  Its blocks are companion matrices of a
divisibility chain
\[
  d_1 \mid d_2 \mid \cdots \mid d_n,
\]
where unit entries are allowed and contribute zero-dimensional blocks.  The
same data presents the vector space as a module over the polynomial ring:
\[
  V_A \cong
  \prod_{i=1}^{n} \Q[X]/(d_i).
\]
This is the invariant-factor classification and companion-block construction
of rational canonical form \citep[Chapter VI]{macduffee1943rcf}.  It makes
Smith normal form the natural computational and proof-theoretic interface.

A formal construction has several obligations hidden by a textbook proof.
The matrix $XI-A$ must use a fixed sign and index convention.  The executable
polynomial representation must agree with mathlib's standard polynomial
type.  Unit factors and dimension zero require explicit semantics.  The sum
of factor degrees must supply an actual equivalence between dependent block
indices and the original finite index.  Finally, a similarity result should
carry an inverse matrix rather than appeal to a determinant or an unspecified
existence proof.

We implement these obligations in Lean 4~\citep{deMouraUllrich2021Lean4} on top of
mathlib~\citep{mathlib2020}.  The underlying HNF/SNF library supplies a strong
result
\[
  U(XI-A)V=S
\]
together with chosen inverses of $U$ and $V$ and both inverse identities.
The application then proves:

\begin{enumerate}
\item the executable presentation is mathlib's characteristic matrix;
\item its normalized Smith diagonal gives a cyclic module decomposition;
\item the factor product and final factor are the characteristic and minimal
  polynomials, respectively;
\item an explicit basis gives
  $P^{-1}AP=\RCF(A)$, with a two-sided inverse certificate for $P$;
\item normalized factors and $\RCF(A)$ are invariant under an explicit
  similarity certificate and independent of the chosen basis.
\end{enumerate}

Ma, Wang, and Wen also derive a rational-canonical theorem through a general
matrix-decomposition bridge \citep[Sections 5.1 and 6.2]{maWangWen2026}.  Their
result is structured existence built from noncomputable choices; it does not
supply an executable reduction, normalized-factor uniqueness, or the explicit
similarity certificate constructed here.  This states the overlap without
using their work as evidence for this implementation.

The rational-canonical facade, paper, benchmark, and artifact are versioned
independently from the frozen 1.2.2 core.  In particular,
\code{import NormalForms} does not expose this application.
