\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage{microtype}
\usepackage{amsmath,amssymb,amsthm,mathtools}
\usepackage{booktabs}
\usepackage[numbers,sort&compress]{natbib}
\usepackage{hyperref}

\hypersetup{
  colorlinks=true,
  citecolor=blue,
  linkcolor=blue,
  urlcolor=blue
}

\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\Q}{\mathbb{Q}}
\newcommand{\HNF}{\operatorname{HNF}}
\newcommand{\bits}{\operatorname{bits}}
\newcommand{\spanrows}{\operatorname{rowspan}}
\setlength{\emergencystretch}{3em}

\title{A Certified Modular Hermite Normal Form Kernel in Lean 4:\\
Determinant Contracts, Canonicality, and Explicit Bit Budgets}
\author{Junye Ji}
\date{Research version 0.1.0, July 2026}

\begin{document}

\maketitle

\begin{abstract}
Modulo-determinant Hermite normal form algorithms control intermediate
coefficients by reducing arithmetic modulo a legal determinant multiple.  We
give an executable Lean 4 formalization of the Domich--Kannan--Trotter value
schedule for full-column-rank integer matrices.  Its input type records a
positive modulus and divisibility by one explicitly nonsingular selected
minor; a different elementary-divisor contract is represented by a different
type.  The executable kernel returns its candidate, stage schedule, and exact
operation-category telemetry.  We prove that every transition preserves an
augmented row lattice, that the legal modulus removes the augmentation at the
endpoint, and that the resulting full-column Hermite shape is the unique
canonical HNF.  A checked fraction-free rank profile lifts this result to a
total executable wrapper for rectangular and rank-deficient matrices.  The
development also proves a closed operation bound, intermediate and final
coefficient bounds, and a schoolbook binary cost budget.  The extended-gcd
charge mirrors the actual Lean implementation based on
\code{Int.gcdA}/\code{Int.gcdB}; its value is proved equal to the standard
operation rather than replaced by another algorithm.  The modular result uses
the project's existing strong HNF type with an explicit inverse.  Its
transform is transported from the stable core only after candidate equality,
so the modular bit theorem is stated precisely for the value kernel.  A
pinned artifact freezes 157 facade entries, audits 43 theorem roots, and
records raw native observations for three deterministic cases.
\end{abstract}

\section{Introduction}

Hermite normal form is canonical, but elementary row reduction alone does not
control the size of intermediate integers.  Modulo-determinant algorithms
replace uncontrolled representatives by bounded congruent values and recover
the exact canonical matrix from a sufficiently strong modulus.  The schedule
formalized here follows the algorithm attributed to Domich, Kannan, and
Trotter~\citep{domich1987hnf}; the same attribution appears in the FLINT
integer-matrix documentation~\citep{flint2026fmpzmat}.  Our implementation is
pure Lean and has no FLINT runtime or proof dependency.

The formalization must answer several questions that a conventional routine
can leave implicit.  Which modulus theorem is required?  What invariant is
preserved by centered reductions?  Why does a locally reduced candidate equal
the project's precise row-HNF convention?  Does a rectangular or
rank-deficient wrapper preserve the direction of the relation map?  Finally,
does a complexity statement charge the extended gcd actually executed by
Lean, or a more convenient substitute?

The independent facade
\begin{center}
  \code{NormalForms.Research.ModularHNF}
\end{center}
answers these questions without changing the stable \code{NormalForms}
facade.  The main contributions are:
\begin{itemize}
  \item distinct typed contracts for determinant-modulus and
        elementary-divisor modular algorithms;
  \item an executable full-column-rank value kernel, exact stage telemetry,
        and a proof that its legal candidate is the canonical core HNF;
  \item a total general-rank wrapper based on a checked integer-only,
        fraction-free rank profile;
  \item explicit operation and coefficient bounds, including a sharp final
        bound by the legal modulus;
  \item a cost mirror for the exact standard integer XGCD path and a composed
        schoolbook bit-operation budget; and
  \item an independently reproducible API manifest, axiom audit, benchmark,
        manuscript, and container profile.
\end{itemize}

Lean 4 and mathlib provide the ambient matrix, determinant, module, and
normal-form theory~\citep{demoura2021lean4,mathlib2020}.  All executable
branches use the project's constructive integer Euclidean instance.

\section{Contracts and result boundary}

Let $A\in\Z^{m\times n}$ with $n\le m$.  A
\code{FullColumnRankWitness A} stores an embedding
$\rho:\{0,\ldots,n-1\}\hookrightarrow\{0,\ldots,m-1\}$ and proves
\[
  \det A[\rho,*]\ne0.
\]
This evidence yields $n\le m$ and rank $n$ after mapping to $\Q$.  The
determinant-modulus witness then stores $D\in\Z$ with
\[
  0<D,
  \qquad |\det A[\rho,*]|\mid |D|.
\]
The selected minor makes the rectangular contract concrete and checkable.

FLINT also exposes a different routine whose modulus is a multiple of the
largest elementary divisor.  That premise is represented by
\code{ElementaryDivisorModulusWitness}.  The disjoint sum
\code{ModulusWitness} can describe either contract, but the DKT entry
\code{runWithDeterminantWitness} accepts only the determinant witness.  In
particular, an elementary-divisor modulus cannot be passed to the wrong
transition system merely because both are positive integers.

The raw return type contains
\[
  (H_{\mathrm{raw}},D_{\mathrm{final}},\text{stages},\text{operations}).
\]
It is deliberately not a certificate.  The public strong entry
\code{modularHNFFullRank} returns the established
\code{HNFResultFin A}, hence stores $U$, $U^{-1}$, $H$, and proofs
\[
  UA=H,\qquad U^{-1}U=I,\qquad UU^{-1}=I.
\]
The matrix field $H$ is definitionally the independently computed raw
candidate.  Once canonical equality has been proved, the transform fields are
transported from the stable core result.  This design avoids a second result
type but also fixes the cost scope: the modular transition theorem prices the
value kernel, not construction of $U$ and $U^{-1}$.

\section{Executable modular schedule}

The kernel processes columns from left to right.  At column $k$, the canonical
pivot row is $k$.  If the pivot is zero it is seeded by the current modulus.
Each nonzero entry below the pivot is combined with it by an extended gcd.  If
$a$ and $b$ are the pair and
\[
  g=ua+vb,
\]
the corresponding two rows are replaced by the unimodular pair
\[
  \begin{pmatrix}
    u&v\\
    -b/g&a/g
  \end{pmatrix}.
\]
Only the active suffix is stored, and each new suffix entry is represented by
the positive-centered map into $(-D/2,D/2]$.

After the lower entries vanish, a second extended gcd combines the pivot with
the modulus.  Scaling by its B\'ezout coefficient makes the pivot represent
the gcd class.  The modulus is divided by this gcd, the pivot is restored if
its stored representative is zero, and ordinary exact quotient reductions
clear the entries above it.  The event
\code{ModularStageEvent} records the column, modulus before the stage, pivot
gcd, and modulus afterward.

\code{ModularOperationCount} separately records additions, multiplications,
extended-gcd calls, exact divisions, modular reductions, and comparisons.
The counter is constructed in the same branches as the value transition and
is therefore observable on every raw run.  It is not reconstructed from the
final HNF.

\section{Semantic invariant and canonicality}

For a matrix $M$, modulus $D$, and processed prefix length $k$, define
\[
  \mathcal L(M,D,k)=
    \spanrows(M)+
    \left\langle D e_j\mid k\le j<n\right\rangle_{\Z}.
\]
The second summand is the suffix module.  Centered reduction changes a scalar
by an integer multiple of $D$.  Row-pair replacement is unimodular before
reduction.  Pivot scaling is invertible in the relevant residue class because
its coefficient comes from the gcd with $D$.  These facts yield preservation
of $\mathcal L$ across every below-row update, pivot stage, and accumulated
run.

The determinant witness supplies the key endpoint absorption theorem:
\[
  \left\langle D e_j\mid k\le j<n\right\rangle_{\Z}
  \subseteq \spanrows(A).
\]
It follows from the adjugate equations for the selected nonsingular minor and
the divisibility of $D$.  Thus augmentation does not change the input row
lattice.  At $k=n$ the suffix module is already zero, and the raw candidate
has the same row lattice as $A$.

The transition proof simultaneously establishes the full processed Hermite
shape: positive pivots, zeros below them, and canonical residues above.  We
prove uniqueness directly for this convention.  Equality of row lattices
first gives mutual divisibility of corresponding pivots.  Positivity makes
the pivots equal.  Strong induction on columns then turns equality modulo each
pivot plus the canonical residue interval into equality of every entry.  The
resulting theorem is
\[
  \code{runWithDeterminantWitness\_candidate\_eq\_reference}.
\]
No appeal to FLINT, native evaluation, or an unchecked candidate occurs in
this proof.

\section{Total fraction-free wrapper}

A full-column kernel does not directly cover wide or rank-deficient inputs.
The structure \code{FractionFreeRankProfile A} stores a rank $r$, ordered
pivot columns, selected pivot rows, a nonzero $r\times r$ minor $B$, and
integer coefficients $c_{ji}$ satisfying
\[
  \det(B) A_{*,j}
    =\sum_{i<r} c_{ji} A_{*,p_i}.
\]
It additionally requires $c_{ji}=0$ whenever column $j$ occurs before pivot
$p_i$.  This lexicographic support condition is what transports the canonical
leading-zero and pivot-gap behavior, rather than merely proving equality of
column spans.

The runtime function \code{searchRankProfile?} enumerates finite ranks and
row/column functions.  For each candidate it computes Cramer numerators with
the integer adjugate and checks all profile equations.  A completeness proof
uses ordered bases over $\Q$ and a nonzero-row-minor theorem, but this rational
linear algebra is erased from execution.  The promoted result
\code{fractionFreeRankProfile A} contains only integer data and proofs.

The modular kernel runs on the selected pivot columns.  Fraction-free column
identities survive multiplication by its explicit left transform.  The
support condition forces the required zero prefix and tail entries, so the
same transform proves HNF for the complete input.  The total entry
\code{integerHermiteNormalFormModular} covers rank zero, pivot gaps, wide and
tall matrices, and empty dimensions while returning the original
\code{HNFResultFin} type.

\section{Operation and coefficient bounds}

For $m$ rows and $n$ columns, one processed column has the uniform scalar
budget
\[
  B_{\mathrm{stage}}(m,n)
   =12mn+5m+3n+4.
\]
Following the exact recursive target lists and fixed stage charges gives
\[
  \text{operations.total}
    \le n\,B_{\mathrm{stage}}(m,n).
\]
This theorem holds for every raw modulus and does not depend on correctness of
the candidate.

Let $h(M)=\max_{ij}|M_{ij}|$.  Centered modular values and scaled pivot values
are bounded by the live modulus.  An above-pivot update applied to stored
entries bounded by $B$ has the conservative envelope
\[
  G(B)=B+B^2.
\]
Writing $G^t$ for iteration, the observable prefix after $p$ processed columns
obeys
\[
 h(M_p)\le
 G^{m\min(p,n)}\!\left(\max(h(A),|D|)\right).
\]
The bound is intentionally simple and executable; no polynomial claim is made
for this coarse envelope.  Under a legal determinant witness, canonicality
gives the much sharper endpoint result
\[
  |H_{ij}|\le |D|.
\]

\section{Bit-operation model}

The shared reference model represents integers by a sign and a natural binary
magnitude.  It implements carry/borrow addition, schoolbook multiplication,
long division with remainder, and extended gcd, with a natural-number cost
field.  Value projections are proved equal to standard integer operations and
each cost is bounded in operand bit lengths, following the usual schoolbook
model~\citep{brent2010modern}.

The modular kernel calls the project's standard integer XGCD instance, which
is based on \code{Nat.xgcdAux}, \code{Int.gcdA}, and \code{Int.gcdB}.  Reusing
a differently normalized bounded-XGCD implementation would leave a gap between
the cost theorem and executed code.  We instead define
\code{StandardXGCD.natAuxWithCost} with the same remainder sequence and
coefficient updates. Writing $X(a,b)$ for its signed integer wrapper, we prove
\[
  X(a,b).\operatorname{value}
  =\operatorname{xgcd}_{\mathrm{standard}}(a,b).
\]
At each recursive level the mirror charges one long division, two
multiplications, two additions, and control.  Coefficient magnitude is bounded
by
\[
  C(H+1)^t,
\]
where $H$ bounds remainders, $C$ bounds incoming coefficients, and $t$ is the
remaining Euclidean depth.  The verified depth bound
\[
  t\le\bits(a)+\bits(b)+1
\]
yields an operand-height-only XGCD charge.

For a stored height $H$, the modular working width is
\[
  W(H)=2\bits(H)+2,
\]
large enough for a product followed by a linear combination or modular
reduction.  The common scalar charge is the sum of upper charges for one
addition, multiplication, standard XGCD, exact division, modular reduction,
and comparison.  Pricing the exact six-field telemetry gives
\code{ModularOperationCount.bitOperationCost}.  Each primitive cost is proved
below the corresponding common charge, and distributivity yields
\[
  \operatorname{bitCost}(C,H)
  \le C.\operatorname{total}\cdot\operatorname{scalarCharge}(H).
\]
Combining this with the operation theorem and full-run coefficient envelope
produces the theorem
\begin{center}
\code{runRaw\_bitOperationCost\_le}.
\end{center}
The legal witness entry inherits the same bound.  This is a binary cost theorem for the value schedule. It
does not include the stable-core construction of the strong transform after
canonical equality.

\section{Evaluation and reproducibility}

The native executable fixes three small determinant-modulus cases.  Table~1
records deterministic fields from the committed baseline.  The cost columns
show binary lengths of the modeled cost and its formal bound, avoiding large
decimal renderings of conservative bounds.

\begin{table}[htbp]
\centering
\caption{Deterministic modular-kernel corpus.}
\begin{tabular}{lrrrrrr}
\toprule
case & shape & operations & op. bound & height bits & cost bits & bound bits\\
\midrule
scalar & $1\times1$ & 7  & 24  & 6  & 24 & 29\\
tall   & $2\times1$ & 21 & 41  & 14 & 33 & 38\\
square & $2\times2$ & 44 & 136 & 50 & 47 & 52\\
\bottomrule
\end{tabular}
\end{table}

The benchmark executes each case separately and all cases together after one
warmup and for seven measurements.  It records wall time, user/system time,
maximum RSS, complete operation-category counts, input/modulus/result widths,
height-envelope width, and modeled cost widths.  JSON and CSV store every raw
observation plus source, binary, operating-system, CPU, memory, libc, Python,
and Lean fingerprints.  On the captured host all four stages have a 0.02
second median.  These timings are regression evidence only; they are not a
portable pass/fail condition and prove no complexity statement.

The release gate rebuilds the facade, 157-entry API manifest, execution
guards, benchmark, and core-freeze tests; runs the full project tests; emits a
43-root JSON axiom report; validates the committed and fresh baselines; and
builds this manuscript.  The allowed axiom union is exactly
\code{Classical.choice}, \code{Quot.sound}, and \code{propext}.  Native- and
compiler-trust axioms, project axioms, and \code{sorryAx} are rejected.

\section{Trust boundary and limitations}

Lean's kernel checks the semantic, canonicality, operation, coefficient, and
modeled-cost theorems.  Mathlib contributes the registered classical,
quotient, and propositional-extensionality principles visible in the audit.
The exhaustive rank-search code, standard-XGCD mirror, and modular value
kernel are ordinary Lean definitions.  FLINT is neither linked nor invoked by
the artifact.

Three boundaries are deliberate.  First, the full-column determinant-modulus
kernel and the elementary-divisor algorithm have separate contracts; only the
former is implemented here.  Second, the general-rank search is complete but
exhaustive, so it is a verified total wrapper rather than a fast rank
decomposition.  Third, primitive modular transform tracing is absent.  The
strong result's $U$ and $U^{-1}$ are transported from the stable core after
matrix equality; their construction cost is not counted as modular work.
These boundaries prevent a value-kernel theorem from being presented as a
strong-transform complexity result.

\section{Conclusion}

The development turns a modulo-determinant HNF schedule into a typed,
executable, and independently auditable Lean component.  Its legal modulus is
explicit, its candidate is proved canonical through an augmented-lattice
invariant, and its rank-deficient wrapper carries checked fraction-free data.
Operation counts, coefficient envelopes, and schoolbook costs refer to the
executed value path, including the exact standard XGCD implementation.  The
remaining transform-cost boundary is stated rather than hidden, leaving a
clear target for a future primitive modular operation log.

\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
